\bf A note on the instability and degeneracy of deep learning models

### 2017-07-11 writing/paper-note/chapters/instab.Rmd
edits from Dr. Vardeman

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM. The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing this joint specification:
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$
Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well. In particular, the number $N_H$ of hiddens may also potentially and arbitrarily increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$
Again like the RBM case, the DBM model specification examples a FSFS model.

@@

The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to better ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model. Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$. This leads to the following definition.
```{definition, name = "S-unstable FSFS", label="instabFSFS", echo=TRUE}
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
```
In other words, a model is S-unstable if, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$. A FSFS model formulation may be termed S-stable if it fails to be S-unstable.

@@

S-unstable FSFS model sequences may be undesirable for several reasons. One is that small changes in data can lead to overly-sensitive changes in probability. Consider, for example, the quantity given by
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.

@@

For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability. We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which could be potentially small. The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.

>>>>

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_H})$ denote the $N_H$ random variables for hiddens with support $\mathcal{X}^{N_H}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \alpha \in \mathbb{R}^{N_H}$, $\boldsymbol \beta \in \mathbb{R}^N$, and $\Gamma$ as a matrix with dimension $N_H \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ then has the joint probability mass function
$$
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
$$
where 
$$
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
$$ 
is the normalizing function. Let $\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}$ with $q(N) = N + N_H + N*N_H$ denote the vector of parameters for the RBM. The probability mass function for the visible variables $X_1, \dots, X_N$ follows from marginalizing the joint specification
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
$$
Note that the vector of model parameters $\boldsymbol \theta_N$, of size $q(N)$, grows in size as a function of sample dimension $N$ to accommodate the dimension of visible variables $X_1, \dots, X_N$, and one may further choose the number $N_H$ of hidden variables to change with $N$ as well. In particular, the number $N_H$ of hiddens may also potentially and arbitrarily increase with $N$. Additionally, as $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM model specification for visibles is a FSFS model. This example also indicates that models formed by marginalizing a base FSFS model (e.g., a type of exponential family model) is again a FSFS model class.

@@

The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as
$$
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
$$
Again like the RBM case, the DBM model specification is an example of a FSFS model.

@@

The main idea is that, in formulating FSFS models for potentially increasing numbers of variables (i.e., for $N \rightarrow \infty$), the ratio (\@ref(eq:elpr)) should remain bounded to ensure model stability, requiring that the largest probability possible under $P_{\theta_N}$ maintain a fixed order of magnitude relative to the smallest probability allowed under the same model. Specifically, the log of the ratio should grow at mostly linearly with the sample size $N$. This leads to the following definition.
```{definition, name = "S-unstable FSFS", label="instabFSFS", echo=TRUE}
Let $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ be a sequence of FSFS model parameters where the size of the model $q(N)$ is a function of the number of random variables $N$. A FSFS model formulation is \emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of variables increase ($N \rightarrow \infty$), 
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
```
In other words, a model is S-unstable if, given any $C > 0$, there exists an integer $N_C > 0$ so that $\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C$ for all $N \ge N_C$. A FSFS model formulation may be termed S-stable if it fails to be S-unstable.

@@

S-unstable FSFS model sequences may be undesirable for several reasons. One is that small changes in data can lead to very large changes in probability. Consider, for example, the quantity given by
$$
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
$$
which represents the biggest log-probability ratio for a one component change in data outcomes at a FSFS parameter $\boldsymbol \theta_N$. We then have the following (non-asymptotic) result.

@@

For a large class of models that covers a broad range of applications (including "deep learning"), we have developed a formal definition of instability in model probability structure and elucidated multiple consequences of instability. We have shown for FSFS models that instability manifests through small changes in data leading to potentially large changes in probability as well as the potential to place all probability on certain modal subsections of the sample space, which potentially could be small. The FSFS model class is quite broad and, particularly in developing FSFS models for large data sets, some caution should be used in parameter specification to control effects of model instability.

### 2017-07-11 writing/paper-note/docs/note.tex
edits from Dr. Vardeman

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X=(X_1,\ldots,X_N)\) denote the \(N\) random variables for
visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H=(H_1,\ldots,H_{N_H})\) denote the \(N_H\) random
variables for hiddens with support \(\mathcal{X}^{N_H}\) where
\(\mathcal{X} = \{-1,1\}\). For parameters
\(\boldsymbol \alpha \in \mathbb{R}^{N_H}\),
\(\boldsymbol \beta \in \mathbb{R}^N\), and \(\Gamma\) as a matrix with
dimension \(N_H \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) then has the
joint probability mass function \[
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
\] where \[
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
\] is the normalizing function. Let
\(\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
with \(q(N) = N + N_H + N*N_H\) denote the vector of parameters for the
RBM. The probability mass function for the visible variables
\(X_1, \dots, X_N\) follows from marginalizing this joint specification:
\[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
\] Note that the vector of model parameters \(\boldsymbol \theta_N\), of
size \(q(N)\), grows in size as a function of sample dimension \(N\) to
accommodate the dimension of visible variables \(X_1, \dots, X_N\), and
one may further choose the number \(N_H\) of hidden variables to change
with \(N\) as well. In particular, the number \(N_H\) of hiddens may
also potentially and arbitrarily increase with \(N\). Additionally, as
\(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM model specification for
visibles is a FSFS model. This example also indicates that models formed
by marginalizing a base FSFS model (e.g., a type of exponential family
model) is again a FSFS model class.

@@

The probability mass function for the visible random variables
\(X_1, \dots, X_N\) follows from this joint specification as \[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
\] Again like the RBM case, the DBM model specification examples a FSFS
model.

@@

To define or measure instability in FSFS models, it is useful to
consider the behavior of a data model sequence \(P_{\theta_N}\). A
relevant quantity to this end is a (scaled) extremal log-probability
ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eq:elpr}
\end{align}
The main idea is that, in formulating FSFS models for potentially
increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
the ratio (\eqref{eq:elpr}) should remain bounded to better ensure model
stability, requiring that the largest probability possible under
\(P_{\theta_N}\) maintain a fixed order of magnitude relative to the
smallest probability allowed under the same model. Specifically, the log
of the ratio should grow at mostly linearly with the sample size \(N\).
This leads to the following definition.
\BeginKnitrBlock{definition}[S-unstable FSFS]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FSFS) \fi{} }Let
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
model parameters where the size of the model \(q(N)\) is a function of
the number of random variables \(N\). A FSFS model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
variables increase (\(N \rightarrow \infty\)),
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a model is S-unstable if,
given any \(C > 0\), there exists an integer \(N_C > 0\) so that
\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
fails to be S-unstable.

@@

S-unstable FSFS model sequences may be undesirable for several reasons.
One is that small changes in data can lead to overly-sensitive changes
in probability. Consider, for example, the quantity given by \[
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\] which represents the biggest log-probability ratio for a one
component change in data outcomes at a FSFS parameter
\(\boldsymbol \theta_N\). We then have the following (non-asymptotic)
result.

@@

For a large class of models that covers a broad range of applications
(including ``deep learning''), we have developed a formal definition of
instability in model probability structure and elucidated multiple
consequences of instability. We have shown for FSFS models that
instability manifests through small changes in data leading to
potentially large changes in probability as well as the potential to
place all probability on certain modal subsections of the sample space,
which could be potentially small. The FSFS model class is quite broad
and, particularly in developing FSFS models for large data sets, some
caution should be used in parameter specification to control effects of
model instability.

>>>>

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X=(X_1,\ldots,X_N)\) denote the \(N\) random variables for
visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H=(H_1,\ldots,H_{N_H})\) denote the \(N_H\) random
variables for hiddens with support \(\mathcal{X}^{N_H}\) where
\(\mathcal{X} = \{-1,1\}\). For parameters
\(\boldsymbol \alpha \in \mathbb{R}^{N_H}\),
\(\boldsymbol \beta \in \mathbb{R}^N\), and \(\Gamma\) as a matrix with
dimension \(N_H \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) then has the
joint probability mass function \[
P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \qquad \tilde{\boldsymbol x} = (\boldsymbol h, \boldsymbol x) \in \mathcal{X}^{N+N_H},
\] where \[
\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \mathcal{X}^{N + N_H}} \exp\left[ \boldsymbol \alpha^T \boldsymbol h + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^T \Gamma \boldsymbol x\right], \qquad \boldsymbol \theta_N \in \Theta_N,
\] is the normalizing function. Let
\(\boldsymbol \theta_N = (\boldsymbol \alpha, \boldsymbol \beta,\Gamma) \in \Theta_N \subset \mathbb{R}^{q(N)}\)
with \(q(N) = N + N_H + N*N_H\) denote the vector of parameters for the
RBM. The probability mass function for the visible variables
\(X_1, \dots, X_N\) follows from marginalizing the joint specification
\[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \mathcal{X}^{N_H}} P_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \mathcal{X}^N.
\] Note that the vector of model parameters \(\boldsymbol \theta_N\), of
size \(q(N)\), grows in size as a function of sample dimension \(N\) to
accommodate the dimension of visible variables \(X_1, \dots, X_N\), and
one may further choose the number \(N_H\) of hidden variables to change
with \(N\) as well. In particular, the number \(N_H\) of hiddens may
also potentially and arbitrarily increase with \(N\). Additionally, as
\(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM model specification for
visibles is a FSFS model. This example also indicates that models formed
by marginalizing a base FSFS model (e.g., a type of exponential family
model) is again a FSFS model class.

@@

The probability mass function for the visible random variables
\(X_1, \dots, X_N\) follows from this joint specification as \[
P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} P_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}), \qquad \boldsymbol x \in \mathcal{X}^N
\] Again like the RBM case, the DBM model specification is an example of
a FSFS model.

@@

To define or measure instability in FSFS models, it is useful to
consider the behavior of a data model sequence \(P_{\theta_N}\). A
relevant quantity to this end is a (scaled) extremal log-probability
ratio (ELPR)
\begin{align}
\frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] \equiv \frac{1}{N} \text{ELPR}_N(\boldsymbol \theta_N). \label{eq:elpr}
\end{align}
The main idea is that, in formulating FSFS models for potentially
increasing numbers of variables (i.e., for \(N \rightarrow \infty\)),
the ratio (\eqref{eq:elpr}) should remain bounded to ensure model
stability, requiring that the largest probability possible under
\(P_{\theta_N}\) maintain a fixed order of magnitude relative to the
smallest probability allowed under the same model. Specifically, the log
of the ratio should grow at mostly linearly with the sample size \(N\).
This leads to the following definition.
\BeginKnitrBlock{definition}[S-unstable FSFS]
\protect\hypertarget{def:instabFSFS}{}{\label{def:instabFSFS}
\iffalse (S-unstable FSFS) \fi{} }Let
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) be a sequence of FSFS
model parameters where the size of the model \(q(N)\) is a function of
the number of random variables \(N\). A FSFS model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if, as the number of
variables increase (\(N \rightarrow \infty\)),
\begin{align*}
\lim\limits_{N \rightarrow \infty} \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N) \equiv \lim\limits_{N \rightarrow \infty} \frac{1}{N} \log \left[\frac{\max\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}{\min\limits_{(x_1, \dots, x_N) \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(x_1, \dots, x_N)}\right] = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a model is S-unstable if,
given any \(C > 0\), there exists an integer \(N_C > 0\) so that
\(\frac{1}{N}\text{ELPR}_N(\boldsymbol \theta_N) > C\) for all
\(N \ge N_C\). A FSFS model formulation may be termed S-stable if it
fails to be S-unstable.

@@

S-unstable FSFS model sequences may be undesirable for several reasons.
One is that small changes in data can lead to very large changes in
probability. Consider, for example, the quantity given by \[
\Delta(\boldsymbol \theta_N) \equiv \max \left\{\log \frac{P_{\boldsymbol \theta_N}(\boldsymbol x)}{P_{\boldsymbol \theta_N}(\boldsymbol x^*)} : \boldsymbol x \text{ }\& \text{ } \boldsymbol x^* \in \mathcal{X}^N \text{ differ in exactly one component}\right\},
\] which represents the biggest log-probability ratio for a one
component change in data outcomes at a FSFS parameter
\(\boldsymbol \theta_N\). We then have the following (non-asymptotic)
result.

@@

For a large class of models that covers a broad range of applications
(including ``deep learning''), we have developed a formal definition of
instability in model probability structure and elucidated multiple
consequences of instability. We have shown for FSFS models that
instability manifests through small changes in data leading to
potentially large changes in probability as well as the potential to
place all probability on certain modal subsections of the sample space,
which potentially could be small. The FSFS model class is quite broad
and, particularly in developing FSFS models for large data sets, some
caution should be used in parameter specification to control effects of
model instability.
